Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-28 Dail_Mad_ext.tex
updated AIC tables
abstract

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing
hypotheses about density dependence and predicting species' responses
to future environmental change or conservation actions. Accounting for
process variation and observation error in such efforts is necessary
to achieve unbiasedness \hl{[may be too strange sounding a term for
  abstract]}, and doing so is possible using recently
developed state-space models.
%for estimators of population change to be unbiased, and state-space
%models have been developed for this
%models are widely used for these purposes because they allow for
%estimating parameters of classical growth models, such as exponential,
%logistic, and Gompertz, while accounting for demographic stochasticity
%and observation error.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models for spatially-replicated
time-series data recently proposed by \citet[Biometrics]{dail_madsen:2011}.
We expand this class of models to accommodate classical growth
models, zero-inflation, and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the  \textbf{R} package
\texttt{unmarked} or
\textbf{JAGS} is also provided.
%is provided for doing so.
%Furthermore, we describe classical and
%Bayesian methods of parameter estimation, and we present methods of
%forecasting population size under future environmental
%conditions.
%These new developments will
%allow researchers to apply these methods to address important
%questions regarding the factors affecting spatial and temporal
%variation in abundance.
A simulation study indicated that bias was negligible and coverage nominal
and an analysis of North American Breeding Bird Survey is presented
for illustration.
}

@@

% Perhaps stack the species

\begin{sidewaystable}
  \centering
  \small
  \caption{Model selection table for ovenbirds and golden-winged
    warblers in Maryland and Virginia, 1966-2010.  We present model name
    and number, number of parameters (Par.), and difference in Akaike's
    information criterion between each model and the top model of that
    set ($\Delta$AIC).  For each species, the first section compares
    models for initial abundance, the second for detection probability,
    and the third for dynamics.}
  \begin{tabular}[h]{lcclcc}
\hline
A. Ovenbirds	&&&		B. Golden-Winged Warblers \\
\hline
Model	&Par.	&$\Delta$AIC	&Model	&Par.	&$\Delta$AIC \\
A.1. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0
&B.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&0 \\
A.2. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&1262.7
&B.2. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0.4 \\
A.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4
&1264.7	&B.3. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&37.8 \\
A.4. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
&0	&B.4. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)	&5
&0 \\
A.5. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind)	&7	&0.9
&B.5. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&3.5 \\
A.6. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(1st)	&5	&5.0
&B.6. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
&5.9 \\
A.7. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&6.4
&B.7. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind)	&7	&9.3 \\
A.8. NB[$\Lambda$(.)$\alpha$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
&10	&0	&B.8. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)
&5	&0 \\
A.9. NB[$\Lambda$(.)$\alpha$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
&10	&8.4
&B.9. ZIP[$\Lambda$(.)$\psi$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(1st)
&6	&1.7 \\
A.10. NB[$\Lambda$(.)$\alpha$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(wind+1st)
&9	&36.5
&B.10. ZIP[$\Lambda$(.)$\psi$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(1st)
&6	&2.0 \\
A.11. NB[$\Lambda$(.)$\alpha$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st)
&10	&38.6
&B.11. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
&7	&2.6 \\
A.12. NB[$\Lambda$(.)$\alpha$(.)]Gompertz[$r$(.)$K$(.)]$p$(wind+1st)
&9	&190.9
&B.12. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz[$r$(.)$K$(.)]$p$(1st)	&6
&3.0 \\
A.13. NB[$\Lambda$(.)$\alpha$(.)]Ricker[$r$(.)$K$(.)]$p$(wind+1st)
&9	&195.1
&B.13. ZIP[$\Lambda$(.)$\psi$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st)
&7	&3.7 \\
A.14. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
&271.3
&B.14. ZIP[$\Lambda$(.)$\psi$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
&7	&4.2 \\
A.15. NB[$\Lambda$(.)$\alpha$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
&9	&273.7
&B.15. ZIP[$\Lambda$(.)$\psi$(.)]Ricker[$r$(.)$K$(.)]$p$(1st)	&7
&5.2 \\
A.16. NB[$\Lambda$(.)$\alpha$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
&9	&1856.7
&B.16. ZIP[$\Lambda$(.)$\psi$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(1st)
&6	&12.5 \\
\hline
\end{tabular}
\end{sidewaystable}

>>>>

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing
hypotheses about density dependence and predicting species' responses
to future environmental change or conservation actions. Failure to
account for observation error in such models can result in
bias, and thus ecologists have increasingly relied on state-space
models to directly model both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect
detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential, Ricker, etc $\dots$), zero-inflation, and
random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the \textbf{R} package
\texttt{unmarked} or \textbf{JAGS} is also provided.
A simulation study indicated that bias was negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlighted how these methods
can be readily applied to existing data, but it also suggested the
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is unavailable.
}

@@

% Perhaps stack the species
\begin{comment}
\begin{sidewaystable}
  \centering
  \small
  \caption{Model selection table for ovenbirds and golden-winged
    warblers in Maryland and Virginia, 1966-2010.  We present model name
    and number, number of parameters (Par.), and difference in Akaike's
    information criterion between each model and the top model of that
    set ($\Delta$AIC).  For each species, the first section compares
    models for initial abundance, the second for detection probability,
    and the third for dynamics.}
  \begin{tabular}[h]{lcclcc}
\hline
A. Ovenbirds	&&&		B. Golden-Winged Warblers \\
\hline
Model	&Par.	&$\Delta$AIC	&Model	&Par.	&$\Delta$AIC \\
A.1. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0
&B.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&0 \\
A.2. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&1262.7
&B.2. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0.4 \\
A.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4
&1264.7	&B.3. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&37.8 \\
A.4. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
&0	&B.4. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)	&5
&0 \\
A.5. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind)	&7	&0.9
&B.5. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&3.5 \\
A.6. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(1st)	&5	&5.0
&B.6. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
&5.9 \\
A.7. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&6.4
&B.7. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind)	&7	&9.3 \\
A.8. NB[$\Lambda$(.)$\alpha$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
&10	&0	&B.8. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)
&5	&0 \\
A.9. NB[$\Lambda$(.)$\alpha$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
&10	&8.4
&B.9. ZIP[$\Lambda$(.)$\psi$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(1st)
&6	&1.7 \\
A.10. NB[$\Lambda$(.)$\alpha$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(wind+1st)
&9	&36.5
&B.10. ZIP[$\Lambda$(.)$\psi$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(1st)
&6	&2.0 \\
A.11. NB[$\Lambda$(.)$\alpha$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st)
&10	&38.6
&B.11. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
&7	&2.5 \\
A.12. NB[$\Lambda$(.)$\alpha$(.)]Gompertz[$r$(.)$K$(.)]$p$(wind+1st)
&9	&192.8
&B.12. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz[$r$(.)$K$(.)]$p$(1st)	&6
&3.0 \\
A.13. NB[$\Lambda$(.)$\alpha$(.)]Ricker[$r$(.)$K$(.)]$p$(wind+1st)
&9	&195.1
&B.13. ZIP[$\Lambda$(.)$\psi$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st)
&7	&3.7 \\
A.14. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
&271.3
&B.14. ZIP[$\Lambda$(.)$\psi$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
&7	&4.2 \\
A.15. NB[$\Lambda$(.)$\alpha$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
&9	&273.7
&B.15. ZIP[$\Lambda$(.)$\psi$(.)]Ricker[$r$(.)$K$(.)]$p$(1st)	&7
&5.2 \\
A.16. NB[$\Lambda$(.)$\alpha$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
&9	&1856.7
&B.16. ZIP[$\Lambda$(.)$\psi$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(1st)
&6	&12.5 \\
\hline
\end{tabular}
\end{sidewaystable}
\end{comment}

% How about separate tables?  Also updated the Gompertz results.
\begin{table}
  \centering
  \small
  \caption{Model selection table for ovenbirds in Maryland and Virginia,
    1966-2010.  We present model name
    and number, number of parameters (Par.), and difference in Akaike's
    information criterion between each model and the top model of that
    set ($\Delta$AIC).  The first section compares
    models for initial abundance, the second for detection probability,
    and the third for dynamics.}
  \begin{tabular}[h]{lcc}
\hline
Model	&Par.	&$\Delta$AIC	\\
\hline
A. Initial Abundance && \\
A.1. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0\\
A.2. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&1262.7\\
A.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4 &1264.7\\
\hline
B. Detection Probability && \\
B.1. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
&0	\\
B.2. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind)	&7	&0.9\\
B.3. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(1st)	&5	&5.0\\
B.4. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&6.4\\
\hline
C. Dynamics && \\
C.1. NB[$\Lambda$(.)$\alpha$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
&10	&0	\\
C.2. NB[$\Lambda$(.)$\alpha$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(wind+1st)
&10	&8.4 \\
C.3. NB[$\Lambda$(.)$\alpha$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(wind+1st)
&9	&36.5\\
C.4. NB[$\Lambda$(.)$\alpha$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(wind+1st)
&10	&38.6\\
C.5. NB[$\Lambda$(.)$\alpha$(.)]Gompertz[$r$(.)$K$(.)]$p$(wind+1st)
&9	&192.8\\
C.6. NB[$\Lambda$(.)$\alpha$(.)]Ricker[$r$(.)$K$(.)]$p$(wind+1st)
&9	&195.1\\
C.7. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
&271.3\\
C.8. NB[$\Lambda$(.)$\alpha$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
&9	&273.7\\
C.9. NB[$\Lambda$(.)$\alpha$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(wind+1st)
&9	&1856.7\\
\hline
\end{tabular}
\end{table}

\begin{table}
  \centering
  \small
  \caption{Model selection table for golden-winged
    warblers in Maryland and Virginia, 1966-2010.  We present model name
    and number, number of parameters (Par.), and difference in Akaike's
    information criterion between each model and the top model of that
    set ($\Delta$AIC).  The first section compares
    models for initial abundance, the second for detection probability,
    and the third for dynamics.}
  \begin{tabular}[h]{lcc}
\hline
Model	&Par.	&$\Delta$AIC\\
\hline
A. Initial Abundance && \\
A.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&0 \\
A.2. NB[$\Lambda$(.)$\alpha$(.)]Growth[$r$(.)]$p$(.)	&4	&0.4 \\
A.3. P[$\Lambda$(.)]Growth[$r$(.)]$p$(.)	&3	&37.8 \\
\hline
B. Detection Probability && \\
B.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)	&5
&0 \\
B.2. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(.)	&4	&3.5 \\
B.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind+1st)	&8
&5.9 \\
B.4. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(wind)	&7	&9.3 \\
\hline
C. Dynamics && \\
C.1. ZIP[$\Lambda$(.)$\psi$(.)]Growth[$r$(.)]$p$(1st)
&5	&0 \\
C.2. ZIP[$\Lambda$(.)$\psi$(.)]AR[$\gamma$(.)$\omega$(.)]$p$(1st)
&6	&1.7 \\
C.3. ZIP[$\Lambda$(.)$\psi$(.)]Growth+Imm.[$r$(.)$\iota$(.)]$p$(1st)
&6	&2.0 \\
C.4. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
&7	&2.5 \\
C.5. ZIP[$\Lambda$(.)$\psi$(.)]Gompertz[$r$(.)$K$(.)]$p$(1st)	&6
&3.0 \\
C.6. ZIP[$\Lambda$(.)$\psi$(.)]AR+Imm.[$\gamma$(.)$\omega$(.)$\iota$(.)]$p$(1st)
&7	&3.7 \\
C.7. ZIP[$\Lambda$(.)$\psi$(.)]Ricker+Imm.[$r$(.)$K$(.)$\iota$(.)]$p$(1st)
&7	&4.2 \\
C.8. ZIP[$\Lambda$(.)$\psi$(.)]Ricker[$r$(.)$K$(.)]$p$(1st)	&7
&5.2 \\
C.9. ZIP[$\Lambda$(.)$\psi$(.)]Constant[$\gamma$(.)$\omega$(.)]$p$(1st)
&6	&12.5 \\
\hline
\end{tabular}
\end{table}
